\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2021-11-28 main.tex
add in examples from vignettes

@@

@@

Matching - Vig 3

@@

\begin{CodeChunk}
\begin{CodeInput}
R> sydmel_climate <- stations %>%
+   rowwise() %>%
+   mutate(ts = list(meteo_pull_monitors(id,
+                                        date_min = "2020-01-01",
+                                        date_max = "2020-12-31",
+                                        var = c("PRCP", "TMAX", "TMIN")) %>%
+                      select(-id))) %>%
+   as_cubble(key = id, index = date, coords = c(long, lat))
\end{CodeInput}
\end{CodeChunk}

>>>>

\hypertarget{support-on-interactive-graphics}{%
\subsection{Support on interactive
graphics}\label{support-on-interactive-graphics}}

@@

This vignette introduces how to perform spatial and temporal manipulate
in a cubble with dplyr verbs. We will illustrate with
\texttt{weatherdata::historical\_tmax} data, which have the historical
maximum temperature recorded for Australian stations with the earliest
dating back to 1859.

\hypertarget{spatial-manipulation}{%
\subsubsection{Spatial manipulation}\label{spatial-manipulation}}

\texttt{historical\_tmax} is already in a nested cubble format, which is
suitable for station-wise manipulation. \texttt{rnoaa} construct the
station id by prefix \texttt{ASN00} to the Bureau of Meteorology (BOM)
station number. In
\href{http://www.bom.gov.au/climate/cdo/about/site-num.shtml}{BOM's
numbering system}, the 2nd and 3rd digit denotes the state a station is
located in and this is equivalent to the 7th to 8th digit in our string.
We can mutate/ filter station in a particular state based on this
information and here we add a column called \texttt{state\_id} and
filter out the ones in New South Wales and Victoria (ranging from 46 to
90).

\begin{CodeChunk}
\begin{CodeInput}
R> tmax_nested <- weatherdata::historical_tmax %>%
+   mutate(state_id = stringr::str_sub(id, 7, 8)) %>%
+   filter(between(state_id, 46, 90))
\end{CodeInput}
\end{CodeChunk}

\hypertarget{temporal-manipulation}{%
\subsubsection{Temporal manipulation}\label{temporal-manipulation}}

There are some operations in the time dimension we would like to make:

\begin{itemize}
\tightlist
\item
  Extract observations in a particular period, say, those in 1971 to
  1975 and 2016 to 2020. This can be used to compare the historical and
  recent climate.
\item
  Summarise daily records into monthly to remove sparsity
\end{itemize}

These time dimension operations can be computed in the long form and
\texttt{stretch()} converts a nested cubble to a long cubble. From a
long cubble, we can write the exact same dplyr codes to complete the two
tasks:

\begin{CodeChunk}
\begin{CodeInput}
R> tmax_long <- tmax_nested %>%
+   stretch() %>%
+   filter(lubridate::year(date) %in% c(1971:1975, 2016:2020)) %>%
+   mutate(
+     month = lubridate::month(date),
+     group = as.factor(ifelse(
+       lubridate::year(date) > 2015,
+       "2016 ~ 2020",
+       "1971 ~ 1975"
+     ))
+   ) %>%
+   group_by(month, group) %>%
+   summarise(tmax = mean(tmax, na.rm = TRUE))
\end{CodeInput}
\end{CodeChunk}

\hypertarget{back-to-spatial}{%
\subsubsection{Back to spatial}\label{back-to-spatial}}

A data quality issue with the \texttt{rnoaa} data is that while it
records the first and last year recorded of each series, it doesn't
report the period of missingness. For example, station
\texttt{ASN00047048} starts it first record in 1957, pauses for a period
from 1963 to 1990, and then resumes it recording till today. Out of the
77 stations in New South Wales and Victoria, 7 stations have this issue
and we would like to remove those stations from the comparison.

Again, this is a station-wise operation and to convert back from the
long cubble to a nested one, use \texttt{tamp()}. Here we keep the
stations with 24 observations (12 months for both periods) after the
monthly aggregation.

\begin{CodeChunk}
\begin{CodeInput}
R> tmax_nested2 <- tmax_long %>%
+   tamp() %>%
+   filter(nrow(ts) == 24)
\end{CodeInput}
\end{CodeChunk}

In some visualisation, we may need information from both spatial and
temporal dimension. One example of this is a glyph map, where spatial
variables, i.e.~\texttt{longitude} and \texttt{latitude}, are used to
construct the major axes and temporal variables, i.e.~\texttt{month} and
\texttt{tmax}, are used to construct the minor axes. This requires these
variables to be in the same table, rather than in different forms. In
cubble, you can append spatial variables that are invariant to the
\texttt{key} withe \texttt{migrate()}

\begin{CodeChunk}
\begin{CodeInput}
R> tmax_long2 <- tmax_nested2 %>%
+   stretch() %>%
+   migrate(latitude, longitude)
\end{CodeInput}
\end{CodeChunk}

\hypertarget{glyph-map}{%
\subsubsection{Glyph map}\label{glyph-map}}

Now its time for the glyph map!

\begin{CodeChunk}
\begin{CodeInput}
R> gly <- glyphs(
+   tmax_long2,
+   x_major = "longitude",
+   y_major = "latitude",
+   x_minor = "month",
+   y_minor = "tmax",
+   height = 0.6,
+   width = 1
+ )
R> 
R> nsw_vic <- ozmaps::abs_ste %>%
+   filter(NAME %in% c("Victoria", "New South Wales"))
R> 
R> plot_map(nsw_vic, fill = "transparent") +
+   geom_path(data = gly, aes(gx, gy,
+     group = interaction(id, group),
+     color = group
+   )) +
+   scale_color_brewer(palette = "Dark2") + 
+   theme_bw() + 
+   coord_sf(xlim = c(140, 155)) + 
+   theme(legend.position = "bottom")
\end{CodeInput}

\begin{center}\includegraphics{figures/unnamed-chunk-10-1} \end{center}

\end{CodeChunk}

\hypertarget{spatial-and-temporal-aggregation}{%
\subsection{Spatial and temporal
aggregation}\label{spatial-and-temporal-aggregation}}

On top of \texttt{aus\_climate} data in this package,
\texttt{climate\_full} from package
\href{https://github.com/huizezhang-sherry/weatherdata}{\texttt{weatherdata}}
has daily climate data of 639 Australia stations from 2016 to 2020. This
is where these stations locate in an Australia map:

This is a lot of stations to look at for one climate variable and they
can't all fit into a glyph map. What we can do is to group stations into
clusters and look at the aggregated series in the glyph map. In this
vignette, I will introduce how to perform spatial (and temporal)
aggregation using hierarchical data structure in cubble.

First let's add a new column \texttt{ll} for calculating distance matrix
in the next section and aggregate the daily precipitation into weekly
measure. These steps should be familiar from the vignette \emph{Data
manipulation with cubble}.

\begin{CodeChunk}
\begin{CodeInput}
R> station_nested <- weatherdata::climate_full %>%
+   mutate(ll = s2::s2_lnglat(long, lat)) %>%
+   stretch() %>%
+   mutate(wk = lubridate::week(date)) %>%
+   group_by(wk) %>%
+   summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
+   tamp()
\end{CodeInput}
\end{CodeChunk}

\hypertarget{hierarchical-data-structure}{%
\subsubsection{Hierarchical data
structure}\label{hierarchical-data-structure}}

Imposing a clustering structure can be thought of as building a
hierarchical structure where stations are nested within clusters. As an
example to illustrate here, we use a kmean clustering algorithm based on
the distance matrix and specify the number of centers to be 20. More
complex algorithms can also be used for more complex problem, as long as
a mapping from each station id to the cluster id can be constructed. We
then join this data to our station data:

\begin{CodeChunk}
\begin{CodeInput}
R> dist_raw <- scale(s2::s2_distance_matrix(station_nested$ll, station_nested$ll))
R> 
R> cluster_res <- tibble(
+   id = station_nested$id,
+   cluster = kmeans(dist_raw, centers = 20, nstart = 50)$cluster
+ )
R> 
R> station_nested <- station_nested %>%
+   left_join(cluster_res)
\end{CodeInput}
\end{CodeChunk}

One thing we hope to do with the cluster is to find the coordinates of
the centroid. These are variables variant to the station but invariant
to the cluster and it would be nice to have a function that structure
each cluster as a row. \texttt{switch\_key()} is the function that does
this: it lets you to specify a new key, say \texttt{cluster} and nests
all spatial variables variant to \texttt{cluster} into a column.
Temporal observations from different stations while within the same
cluster are bound in the nested column \texttt{ts}.

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_nested <- station_nested %>%
+   switch_key(cluster)
\end{CodeInput}
\end{CodeChunk}

This structure makes it easy to compute cluster level variable, although
there are a few steps to calculate the centroid coordinates: we need to
find the convex hull that wraps around the cluster, make it a polygon,
find the centroid of the polygon and finally, extract the x and y
coordinate of each centroid:

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_nested <- cluster_nested %>%
+   mutate(
+     chull = list(chull(.val$long, .val$lat)),
+     ll_cluster = sf::st_as_sfc(
+       s2::s2_make_polygon(c(.val$long[chull]),
+         c(.val$lat[chull]),
+         oriented = FALSE
+       )
+     ),
+     centroid = s2::s2_centroid(ll_cluster),
+     cent_long = s2::s2_x(centroid),
+     cent_lat = s2::s2_y(centroid)
+   )
\end{CodeInput}
\end{CodeChunk}

After we have got \texttt{cluster\_nested}, spatial and temporal data at
both levels can be easily obtained. If we use \texttt{station} and
\texttt{cluster} prefix to denote the two levels and \texttt{nested} and
\texttt{long} for whether the data shows the spatial or temporal
dimension, the relationship among the four datasets can be illustrated
in the following workflow:

Start with the original \texttt{station\_nested}, \texttt{stretch()}
expands the \texttt{ts} column with each station (\texttt{id}) forming a
group and attach variables invariant to \texttt{id} as an attribute.
\texttt{switch\_key()} changes the \texttt{key} from \texttt{id} to
\texttt{cluster} and nests all the spatial variables that variant to
\texttt{cluster}. \texttt{stretch()} \texttt{cluster\_nested} will store
variables that are invariant to \texttt{cluster} as a tibble in the
attribute.

Now we can obtain the aggregated series as described in the workflow
diagram above and construct the glyph map with
\texttt{GGally::glyphs()}:

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_long <- cluster_nested %>%
+   stretch(ts) %>%
+   group_by(wk) %>%
+   summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
+   migrate(cent_long, cent_lat)
R> 
R> gly <- GGally::glyphs(cluster_long,
+   x_major = "cent_long", x_minor = "wk",
+   y_major = "cent_lat", y_minor = "prcp",
+   height = 2, width = 4
+ )
R> 
R> state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
R> plot_map(state_map) +
+   geom_text(
+     data = cluster_nested,
+     aes(x = cent_long, y = cent_lat, label = cluster)
+   ) +
+   geom_path(
+     data = gly,
+     aes(x = gx, y = gy, group = gid)
+   )
\end{CodeInput}

\begin{center}\includegraphics{figures/unnamed-chunk-15-1} \end{center}

\end{CodeChunk}

We can also look at the precipitation of each individual station within
the same cluster:

\begin{CodeChunk}
\begin{CodeInput}
R> station_long <- station_nested %>%
+   stretch(ts) %>%
+   migrate(cluster)
R> station_long %>%
+   ggplot(aes(x = wk, y = prcp, group = id)) +
+   geom_line(alpha = .3) +
+   facet_wrap(vars(cluster), scales = "free_y", ncol = 4) +
+   theme_bw()
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{figures/unnamed-chunk-16-1}

\caption[shtishisa]{shtishisa}\label{fig:unnamed-chunk-16}
\end{figure}
\end{CodeChunk}

Lastly, there is one series on Tasmania island standing out from others,
lets look at where it is:

\begin{CodeChunk}
\begin{CodeInput}
R> # this part still needs some fixing
R> tas_latlong <- station_nested %>%
+   filter(lat < -40) %>%
+   mutate(p = max(ts$prcp)) %>%
+   strip_rowwise() %>%
+   filter(p == max(p))
R> 
R> state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
R> plot_map(state_map) +
+   geom_point(data = station_nested, aes(x = long, y = lat), size = 0.5) +
+   geom_sf(data = cluster_nested, aes(geometry = ll_cluster), fill = "transparent") +
+   geom_point(data = tas_latlong, aes(x = long, y = lat), col = "red")
\end{CodeInput}

\begin{center}\includegraphics{figures/unnamed-chunk-17-1} \end{center}

\end{CodeChunk}

@@

The water level data comes from
\href{http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full}{Bureau
of Meteorology} and has a copy in \texttt{weatherdata}. Here we extract
the water course level and add a column annotate this data of type
\texttt{river}. For the rainfall data, we will still use the
\texttt{weatherdata::climate\_full}, filtering for Victorian stations in
2020 should be pretty familiar by now. Again, we first look at where
these stations are on the map first:

\begin{CodeChunk}
\begin{CodeInput}
R> river <- weatherdata::water %>%
+   stretch() %>%
+   select(date, Water_course_level) %>%
+   tamp() %>%
+   mutate(type = "river")
R> 
R> climate <- weatherdata::climate_full %>%
+   filter(between(stringr::str_sub(id, 7, 8), 76, 90)) %>%
+   stretch() %>%
+   filter(lubridate::year(date) == 2020) %>%
+   tamp() %>%
+   mutate(type = "climate")
R> 
R> vic_map <- rmapshaper::ms_simplify(ozmaps::abs_ste %>% filter(NAME == "Victoria"))
R> plot_map(vic_map) +
+   geom_point(
+     data = dplyr::bind_rows(river, climate),
+     aes(x = long, y = lat, color = type)
+   ) +
+   scale_color_brewer(palette = "Dark2") +
+   theme_bw()
\end{CodeInput}

\begin{center}\includegraphics{figures/unnamed-chunk-18-1} \end{center}

\end{CodeChunk}

\hypertarget{theory}{%
\subsubsection{Theory}\label{theory}}

Temporal matching checks how spatially matched pairs align temporally.
We use the following chart to illustrate how the temporal matching
works:

{\centering \includegraphics{figures/unnamed-chunk-19-1}

\caption[shtishisaasdf]{shtishisaasdf}\label{fig:unnamed-chunk-19}
\end{figure}
\end{CodeChunk}

For each spatially matched pair, say \texttt{A} and \texttt{a}, we first
find the largest \texttt{n} points in each series, colored in brown
points here. Here we use the largest three but you can tune this number
by \texttt{temporal\_n\_highest}. Then we construct the interval of the
largest points from one series and see how many points, from the other
series, fall into the intervals. The series used to construct the
interval is controlled by \texttt{temporal\_independent} and the window
size by \texttt{temporal\_window} with a default of 5.

In this illustration, we construct the interval based on series
\texttt{A} and two of the three peaks from \texttt{a} falls into this
interval at Time 7 and 27.

\hypertarget{rainfall-translates-into-river-level}{%
\subsubsection{Rainfall translates into river
level}\label{rainfall-translates-into-river-level}}

There's another mandatory argument that hasn't been introduced above:
\texttt{temporal\_var\_to\_match}. This argument controls the variable
to match and it needs to appear in both the \texttt{major} and
\texttt{minor} set. In the water level matching example, we match the
variable \texttt{Water\_course\_level} from \texttt{river} to
\texttt{prcp} from \texttt{climate}, hence need to manually rename one
of them to match the other, here we rename \texttt{Water\_course\_level}
to \texttt{prcp} in \texttt{river}:

\begin{CodeChunk}
\begin{CodeInput}
R> river <- river %>%
+   stretch() %>%
+   rename(prcp = Water_course_level) %>%
+   tamp()
\end{CodeInput}
\end{CodeChunk}

Now we use \texttt{match\_sites()} to first pair the weather stations
with the river gauges spatially and then apply the temporal matching on
\texttt{prcp}. We will construct the interval based on peaks in
\texttt{climate} since we would expect a lag effect for precipitation to
flow into the river and cause a raise in river level, hence
\texttt{temporal\_independent\ =\ climate}. We select the 30 highest
peak from the series to construct the match by setting
\texttt{temporal\_n\_highest\ =\ 30}. This is a tuning parameter and you
can start with 10\% of the points of one series (here we have daily data
for a year, 10\% is roughly 30 points). \texttt{temporal\_min\_match}
filters out pairs don't have enough match and to return all the pairs,
set \texttt{temporal\_min\_match} to \texttt{0}.

\begin{CodeChunk}
\begin{CodeInput}
R> res <- match_sites(river, climate,
+   temporal_var_to_match = prcp,
+   temporal_independent = climate,  
+   temporal_n_highest = 30,
+   temporal_min_match = 15
+ )
R> 
R> res
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], prcp [dbl]
  id          name                lat  long type  ts        .dist .group n_match
  <chr>       <chr>             <dbl> <dbl> <chr> <list>    <dbl>  <int>   <int>
1 405234      SEVEN CREEKS @ D~ -36.9  146. river <tibble ~  6.15      5      21
2 ASN00082042 strathbogie       -36.8  146. clim~ <tibble ~  6.15      5      21
3 404207      HOLLAND CREEK @ ~ -36.6  146. river <tibble ~  8.54     10      21
4 ASN00082170 benalla airport   -36.6  146. clim~ <tibble ~  8.54     10      21
5 230200      MARIBYRNONG RIVE~ -37.7  145. river <tibble ~  6.17      6      19
6 ASN00086038 essendon airport  -37.7  145. clim~ <tibble ~  6.17      6      19
7 406213      CAMPASPE RIVER @~ -37.0  145. river <tibble ~  1.84      1      18
8 ASN00088051 redesdale         -37.0  145. clim~ <tibble ~  1.84      1      18
\end{CodeOutput}
\end{CodeChunk}

The output from temporal matching is also a cubble, with additional
column \texttt{.dist} and \texttt{.group} inherent from spatial matching
and \texttt{n\_match} for the number of matched temporal peaks. Then you
can use this output to plot the location of match or to look at the
series:

\begin{CodeChunk}
\begin{CodeInput}
R> p1 <- plot_map(vic_map) +
+   geom_point(
+     data = res,
+     aes(x = long, y = lat, color = type)
+   ) +
+   ggrepel::geom_label_repel(
+     data = res %>% filter(type == "river"),
+     aes(x = long, y = lat, label = .group)
+   ) +
+   scale_color_brewer(palette = "Dark2") +
+   ggtitle("Victoria") +
+   theme_minimal() +
+   theme(
+     panel.grid.major = element_blank(),
+     panel.grid.minor = element_blank(),
+     legend.position = "bottom"
+   )
R> 
R> res_long <- res %>%
+   stretch(ts) %>%
+   migrate(.group, type) %>%
+   mutate(prcp = scale(prcp)[, 1]) %>%
+   switch_key(.group) %>%
+   migrate(type)
R> 
R> p2 <- res_long %>%
+   ggplot(aes(
+     x = date, y = prcp,
+     group = id, color = type
+   )) +
+   geom_line() +
+   facet_wrap(vars(.group)) +
+   scale_color_brewer(palette = "Dark2", guide = "none") +
+   theme_bw() +
+   labs(x = "week") +
+   scale_x_date(date_labels = "%b")
R> 
R> p1 | p2
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/unnamed-chunk-22-1}

\caption[..]{...}\label{fig:unnamed-chunk-22}
\end{figure}
\end{CodeChunk}

\hypertarget{interative-graphic-with-cubble}{%
\subsection{Interative graphic with
cubble}\label{interative-graphic-with-cubble}}

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(123)
R> tmax_data <- weatherdata::climate_full %>%
+   slice_sample(n = 50) %>%
+   stretch() %>%
+   mutate(ym = tsibble::yearmonth(date)) %>%
+   group_by(ym) %>%
+   summarise(
+     tmax = mean(tmax, na.rm = TRUE),
+     tmin = mean(tmin, na.rm = TRUE),
+     prcp = sum(prcp, na.rm = TRUE)
+   ) %>%
+   ungroup(ym) %>%
+   mutate(
+     year = as.factor(year(ym)),
+     month = as.factor(month(ym)),
+     prcp = sqrt(prcp + 0.001)
+   ) %>%
+   tamp() %>%
+   mutate(median_tmax = median(ts$tmax, na.rm = TRUE)) %>%
+   stretch() %>%
+   migrate(median_tmax, long, lat, name) %>%
+   highlight_key(~id)
R> 
R> p1 <- tmax_data %>%
+   ggplot(aes(
+     x = month, y = tmax, group = interaction(year, id),
+     color = median_tmax, label = name
+   )) +
+   geom_line() +
+   facet_wrap(vars(fct_reorder(name, -median_tmax))) +
+   theme(legend.position = "none")
R> 
R> state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
R> p2 <- tmax_data %>%
+   ggplot() +
+   geom_sf(data = state_map, aes(geometry = geometry)) +
+   geom_point(aes(x = long, y = lat, color = median_tmax, label = name)) +
+   theme_void()
R> 
R> bscols(
+   ggplotly(p1, width = 800, height = 800, tooltip = "label") %>%
+     highlight(on = "plotly_selected", off = "plotly_deselect", color = "red"),
+   ggplotly(p2, width = 800, height = 800, tooltip = "label") %>%
+     highlight(on = "plotly_selected", off = "plotly_deselect", color = "red")
+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> sydmel_climate <- stations %>%
+   rowwise() %>%
+   mutate(ts = list(meteo_pull_monitors(id,
+     date_min = "2020-01-01",
+     date_max = "2020-12-31",
+     var = c("PRCP", "TMAX", "TMIN")
+   ) %>%
+     select(-id))) %>%
+   as_cubble(key = id, index = date, coords = c(long, lat))
\end{CodeInput}
\end{CodeChunk}
